Probabilistic Forecasting of Geomagnetic Indices using Gaussian Process Models

### 2017-04-18 chap.tex
Completing Workflow section

\usepackage{amsmath}
\usepackage{txfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{url}
\usepackage{natbib}
\bibliographystyle{biblio.bst}

@@

@@

@@

@@

>>>>

\usepackage{amsmath}
\usepackage{txfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{url}
\usepackage{natbib}
\usepackage{rotating}
\usepackage{adjustbox}
\graphicspath{{figures/}}
\bibliographystyle{biblio.bst}

@@

We selected OMNI data sections 00:00 January 3 2010 to 23:00 January 23 2010 and 20:00 August 5 2011 to 22:00 August 6 2011 for training the GP-AR and GP-ARX models. The first training data section consists of ambient fluctuations of $Dst$ while the second contains a geomagnetic storm. A set of 24 storm events listed in Table \ref{table:validationstorms} is kept aside as a validation data set for selecting model orders $p, p_v, p_b$ while a list of 63 storm events listed in table \ref{table:teststorms} is used for final model evaluation.

@@

We choose a kernel $K_{osa}$ of the form given in equation \ref{eq:usedKernel}. In order to find appropriate values of the hyper-parameters of $K_{osa}$, we apply \emph{grid search}, \emph{coupled simulated annealing} and \emph{maximum likelihood} methods. We fix the parameters $d$ and $\sigma^2$ of $K_{st}$ and model noise to values $0.01$ and $0.2$ respectively, the remaining parameters $w$ and $b$ are kept free to be calculated by model selection. Table \ref{table:modelselection} summarizes the settings used to run each model selection procedure.

\item \textbf{Choice of Performance Metrics}: To compare the performance of GP and GP-ARX models of varying auto-regressive orders, their predictive performance during the validation (and later final evaluation) phase must be calculated using some clearly defined metrics. We use the following as performance metrics for our modeling experiments.

\begin{enumerate}
    \item The mean absolute error.
    \begin{equation}
        MAE = \sum_{t=1}^{n} \left |(Dst(t) - \hat{D}st(t)) \right | / n
    \end{equation}
    \item The root mean square error.
    \begin{equation}
        RMSE = \sqrt{\sum_{t=1}^{n} (Dst(t) - \hat{D}st(t))^2 / n}
    \end{equation}
    \item Correlation coefficient between the predicted and actual value of $Dst$.
    \begin{equation}
        CC = Cov(Dst, \hat{D}st)/\sqrt{Var(Dst) Var(\hat{D}st)}
    \end{equation}
\end{enumerate}

@@

\item \textbf{Final Evaluation}: After selecting GP-AR and GP-ARX models with particular values of auto-regressive orders and kernel hyper-parameters in the model selection phase, the best performing GP-AR and GP-ARX models are then evaluated on the test data set in order to get an unbiased picture of their predictive performance. In our case we use the already chosen test set outlined in table \ref{table:teststorms} for final model evaluation. The GP based OSA prediction models are compared against the naive \emph{Persistence} model which serves as a performance baseline for the OSA prediction of $Dst$.

@@

\section{Experiments \& Results}

\begin{table}[h]
\centering
\caption{Settings of model selection procedures}
\begin{tabular}{l c c c}
\hline
Procedure & Grid Size & Step & Max Iterations \\
\hline
Grid Search & 10 & 0.2 & NA \\
Coupled Simulated Annealing & 4 & 0.2 & 30 \\
Maximum likelihood & NA & 0.2 & 150\\
\end{tabular}
\label{table:modelselection}
\end{table}

\begin{table}[h]
\centering
\caption{Storm events used for model selection of GP-AR and GP-ARX}
\label{table:validationstorms}
\begin{tabular}{llllll}
\hline
Event Id & Start Date & Start Hour & End Date & End Hour & min. Dst \\ \hline
1 & 1995/03/26 & 0500 & 1995/03/26 & 2300 & −107 \\
2 & 1995/04/07 & 1300 & 1995/04/09 & 0900 & −149 \\
3 & 1995/09/27 & 0100 & 1995/09/28 & 0400 & −108 \\
4 & 1995/10/18 & 1300 & 1995/10/19 & 1400 & −127 \\
5 & 1996/10/22 & 2200 & 1996/10/23 & 1100 & −105 \\
6 & 1997/04/21 & 1000 & 1997/04/22 & 0900 & −107 \\
7 & 1997/05/15 & 0300 & 1997/05/16 & 0000 & −115 \\
8 & 1997/10/10 & 1800 & 1997/10/11 & 1900 & −130 \\
9 & 1997/11/07 & 0000 & 1997/11/07 & 1800 & −110 \\
10 & 1997/11/22 & 2100 & 1997/11/24 & 0400 & −108 \\
11 & 2005/06/12 & 1700 & 2005/06/13 & 1900 & −106 \\
12 & 2005/08/31 & 1200 & 2005/09/01 & 1200 & −122 \\
13 & 2006/12/14 & 2100 & 2006/12/16 & 0300 & −162 \\
14 & 2011/09/26 & 1400 & 2011/09/27 & 1200 & −101 \\
15 & 2011/10/24 & 2000 & 2011/10/25 & 1400 & −132 \\
16 & 2012/03/08 & 1200 & 2012/03/10 & 1600 & −131 \\
17 & 2012/04/23 & 1100 & 2012/04/24 & 1300 & −108 \\
18 & 2012/07/15 & 0100 & 2012/07/16 & 2300 & −127 \\
19 & 2012/09/30 & 1300 & 2012/10/01 & 1800 & −119 \\
20 & 2012/10/08 & 0200 & 2012/10/09 & 1700 & −105 \\
21 & 2012/11/13 & 1800 & 2012/11/14 & 1800 & −108 \\
22 & 2013/03/17 & 0700 & 2013/03/18 & 1000 & −132 \\
23 & 2013/05/31 & 1800 & 2013/06/01 & 2000 & −119 \\
24 & 2014/02/18 & 1500 & 2014/02/19 & 1600 & −112
\end{tabular}
\end{table}

\begin{table}[h]
\fontsize{6}{7.5}\selectfont
\centering
\caption{Storm events used to evaluate GP-AR and GP-ARX models}
\label{table:teststorms}
\begin{tabular}{cccccc}
\hline
Event Id & Start Date & Start Time & End Date & End Time & min. Dst \\ \hline
1 & 1998/02/17 & 1200 & 1998/02/18 & 1000 & -100 \\
2 & 1998/03/10 & 1100 & 1998/03/11 & 1800 & -116 \\
3 & 1998/05/04 & 0200 & 1998/05/05 & 0200 & -205 \\
4 & 1998/08/26 & 1000 & 1998/08/29 & 0700 & -155 \\
5 & 1998/09/25 & 0100 & 1998/09/26 & 0000 & -207 \\
6 & 1998/10/19 & 0500 & 1998/10/20 & 0800 & -112 \\
7 & 1998/11/09 & 0300 & 1998/11/10 & 1600 & -142 \\
8 & 1998/11/13 & 0000 & 1998/11/15 & 0400 & -131 \\
9 & 1999/01/13 & 1600 & 1999/01/14 & 2000 & -112 \\
10 & 1999/02/18 & 0300 & 1999/02/19 & 2100 & -123 \\
11 & 1999/09/22 & 2000 & 1999/09/23 & 2300 & -173 \\
12 & 1999/10/22 & 0000 & 1999/10/23 & 1400 & -237 \\
13 & 2000/02/12 & 0500 & 2000/02/13 & 1500 & -133 \\
14 & 2000/04/06 & 1700 & 2000/04/08 & 0900 & -288 \\
15 & 2000/05/24 & 0100 & 2000/05/25 & 2000 & -147 \\
16 & 2000/08/10 & 2000 & 2000/08/11 & 1800 & -106 \\
17 & 2000/08/12 & 0200 & 2000/08/13 & 1700 & -235 \\
18 & 2000/10/13 & 0200 & 2000/10/14 & 2300 & -107 \\
19 & 2000/10/28 & 2000 & 2000/10/29 & 2000 & -127 \\
20 & 2000/11/06 & 1300 & 2000/11/07 & 1800 & -159 \\
21 & 2000/11/28 & 1800 & 2000/11/29 & 2300 & -119 \\
22 & 2001/03/19 & 1500 & 2001/03/21 & 2300 & -149 \\
23 & 2001/03/31 & 0400 & 2001/04/01 & 2100 & -387 \\
24 & 2001/04/11 & 1600 & 2001/04/13 & 0700 & -271 \\
25 & 2001/04/18 & 0100 & 2001/04/18 & 1300 & -114 \\
26 & 2001/04/22 & 0200 & 2001/04/23 & 1500 & -102 \\
27 & 2001/08/17 & 1600 & 2001/08/18 & 1600 & -105 \\
28 & 2001/09/30 & 2300 & 2001/10/02 & 0000 & -148 \\
29 & 2001/10/21 & 1700 & 2001/10/24 & 1100 & -187 \\
30 & 2001/10/28 & 0300 & 2001/10/29 & 2200 & -157 \\
31 & 2002/03/23 & 1400 & 2002/03/25 & 0500 & -100 \\
32 & 2002/04/17 & 1100 & 2002/04/19 & 0200 & -127 \\
33 & 2002/04/19 & 0900 & 2002/04/21 & 0600 & -149 \\
34 & 2002/05/11 & 1000 & 2002/05/12 & 1600 & -110 \\
35 & 2002/05/23 & 1200 & 2002/05/24 & 2300 & -109 \\
36 & 2002/08/01 & 2300 & 2002/08/02 & 0900 & -102 \\
37 & 2002/09/04 & 0100 & 2002/09/05 & 0000 & -109 \\
38 & 2002/09/07 & 1400 & 2002/09/08 & 2000 & -181 \\
39 & 2002/10/01 & 0600 & 2002/10/03 & 0800 & -176 \\
40 & 2002/10/03 & 1000 & 2002/10/04 & 1800 & -146 \\
41 & 2002/11/20 & 1600 & 2002/11/22 & 0600 & -128 \\
42 & 2003/05/29 & 2000 & 2003/05/30 & 1000 & -144 \\
43 & 2003/06/17 & 1900 & 2003/06/19 & 0300 & -141 \\
44 & 2003/07/11 & 1500 & 2003/07/12 & 1600 & -105 \\
45 & 2003/08/17 & 1800 & 2003/08/19 & 1100 & -148 \\
46 & 2003/11/20 & 1200 & 2003/11/22 & 0000 & -422 \\
47 & 2004/01/22 & 0300 & 2004/01/24 & 0000 & -149 \\
48 & 2004/02/11 & 1000 & 2004/02/12 & 0000 & -105 \\
49 & 2004/04/03 & 1400 & 2004/04/04 & 0800 & -112 \\
50 & 2004/07/22 & 2000 & 2004/07/23 & 2000 & -101 \\
51 & 2004/07/24 & 2100 & 2004/07/26 & 1700 & -148 \\
52 & 2004/07/26 & 2200 & 2004/07/30 & 0500 & -197 \\
53 & 2004/08/30 & 0500 & 2004/08/31 & 2100 & -126 \\
54 & 2004/11/07 & 2100 & 2004/11/08 & 2100 & -373 \\
55 & 2004/11/09 & 1100 & 2004/11/11 & 0900 & -289 \\
56 & 2004/11/11 & 2200 & 2004/11/13 & 1300 & -109 \\
57 & 2005/01/21 & 1800 & 2005/01/23 & 0500 & -105 \\
58 & 2005/05/07 & 2000 & 2005/05/09 & 1000 & -127 \\
59 & 2005/05/29 & 2200 & 2005/05/31 & 0800 & -138 \\
60 & 2005/06/12 & 1700 & 2005/06/13 & 1900 & -106 \\
61 & 2005/08/31 & 1200 & 2005/09/01 & 1200 & -131 \\
62 & 2006/04/13 & 2000 & 2006/04/14 & 2300 & -111 \\
63 & 2006/12/14 & 2100 & 2006/12/16 & 0300 & -147 \\ \hline
\end{tabular}%
\end{table}

\begin{table}[h]
\centering
\caption{Evaluation results for models on storm events listed in table \ref{table:teststorms}}
\label{table:results}
\begin{tabular}{l c c c}
\hline
Model & Mean Absolute Error & Root Mean Square Error & Coefficient of Correlation\\ \hline
GP-ARX & 7.252 & 11.93 & 0.972\\
GP-AR & 8.37 & 14.04 & 0.963\\
Persistence & 9.182 & 14.94 & 0.957\\
\end{tabular}
\end{table}
